---
title: "Assignment-5"
author: "Don Radcliffe"
date: "2/14/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

I've had issues getting the here() function to work whenever I have a markdown file within an R project, and had my script in one folder and the data in another non-nested folder.  I can make it work if I nest the data folder within my scripts folder.  But I don't think that's a good organization strategy, because it may confuse somebody else trying to use my data.  Any thoughts on how to work this differently?  In these cases I keep going back to Jenny Bryan's Truman option of setwd().
```{r packages and data, echo = FALSE}
library(here)
library(ggplot2)
library(dplyr)

#here('Assignment-5/data')
setwd('C:/ProgramR/fish497/homework5/Assignment-5/data')
trout <- read.csv('siscowet.csv')
head(trout)
```

As a fisherperson, I might be interested in seeing how fish weight varies by depth, so I know what depth to fish for the big trout.  I would also want to know whether the relationship between depth and weight varies by site, to know whether I can consistently fish the same depth across the Upper Peninsula shore of Lake Superior, or if I have to experiment every time I fish a new locale. 

I can see from the following graph that the depth I fish probably wouldn't affect the weight of  
```{r explore}
trout_exploratory <- trout %>%
  ## removing one crazy outlier
  filter(wgt < 5000) %>%
  ggplot(aes(x = pnldep, y = wgt, color = locID)) +
  ## jittered the points because there were a lot of linear patterns
  geom_jitter(alpha = 0.5) +
  theme_bw()
trout_exploratory
```